Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-08-25 ltx/data_paper/DataPaper.tex
edits to discussion

The correlation analysis specifies approximate supports for both the disturbance covariance function and the connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly. The analysis also assumes the sigmoidal firing rate in the active region and as a result  the kernel and disturbance amplitudes cannot be inferred only through the correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S\ref{sec:EstimationAlgorithm} where the approximated supports are used to estimate the gain of the kernel and the field disturbance variance. Knowing the spatial extent of the connectivity kernel greatly improves the speed of the estimation algorithm as the number of basis functions for the kernel decomposition is significantly reduced  compared to the case where the basis functions are arranged across the entire spatial domain. This also improves the uncertainty of the estimates as a lower number of unknown parameters need to be estimated. The approximated support of the disturbance covariance function facilitates the application of the EM algorithm. This way the disturbance covariance matrix can be decomposed into two parts: an unknown scalar and a constant matrix depending only on the inferred spatial support. Therefore, the estimation problem of the disturbance covariance matrix breaks down to the estimation of a single scalar parameter, improving the accuracy and the convergence property of the algorithm.

The additional Taylor approximation step introduced in the E-step of the algorithm resembles the extended Kalman filtering \cite{Jazwinski1970,Chui2009}. The EKF approximates a stationary non-linear dynamic system with a linear non-stationary one using the first order Taylor expansion around the state estimates. This might imply that the extended Kalman smoother (EKS) is a better candidate --- in terms of implementation --- for the state inference in the proposed algorithm. However, the EKS requires re-linearising about the backward estimates which in turn necessitates the non-linear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation. In addition, the URTSS provides better state estimations compared to the EKS \cite{Haykin2001} and therefore the constructed lower bound will be more accurate, resulting in a better parameter estimation.

@@

Since the non-linearity is a sigmoid function first-order Taylor linearisation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated non-linear behaviour higher order terms might need to be included, this however introduces intractable computational complexity.
\section{Conclusion}
In this chapter, a method has been developed to estimate a spatially continuous neural field model from electrophysiological data, where the kernel spatial support and the field disturbance characteristics were unknown to the estimator.  It was shown that by choosing the synaptic parameter and the observation noise variance within a plausible range these unknown quantities can be identified. This improves the complexity and the accuracy of the EM algorithm by reducing the number of unknown variables in the estimation process. In addition to the connectivity kernel weights and the synaptic parameter, the observation noise and disturbance variances were also identified. The empirical distributions of the estimated non-kernel parameters showed small bias with very small standard deviations.

>>>>

This paper is an extension of previous work in developing a framework for subject-specific neural field modelling. The extensions of the previous work include a method for estimating the support of the connectivity kernel and disturbance covariance, and the development of an EM framework for the neural field equations using the URTSS. The EM framework combined with the estimation of the support of the disturbance enables estimation of the temporal disturbance and noise characteristics of the system. Furthermore, by estimating the support of the connectivity kernel we are able to relax the strong assumption of isotropy in the estimated connectivity structure.

\subsection{Correlation Analysis}
The correlation analysis specifies an approximation for the support of both the disturbance covariance function and the connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly. The analysis also assumes the sigmoidal firing rate in the active region and as a result  the kernel and disturbance amplitudes cannot be inferred only through the correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S\ref{sec:EstimationAlgorithm} where the approximated supports are used to estimate the gain of the kernel and the field disturbance variance. Knowing the spatial extent of the connectivity kernel greatly improves the speed of the estimation algorithm as the number of basis functions for the kernel decomposition is significantly reduced  compared to the case where the basis functions are arranged across the entire spatial domain. This also improves the uncertainty of the estimates as a lower number of unknown parameters need to be estimated. The approximated support of the disturbance covariance function facilitates the application of the EM algorithm. This way the disturbance covariance matrix can be decomposed into two parts: an unknown scalar and a constant matrix depending only on the inferred spatial support. Therefore, the estimation problem of the disturbance covariance matrix breaks down to the estimation of a single scalar parameter, improving the accuracy and the convergence property of the algorithm.

\subsection{Development of the EM for the URTSS}
The additional Taylor approximation step introduced in the E-step of the algorithm resembles the extended Kalman filtering \cite{Jazwinski1970,Chui2009}. The EKF approximates a stationary non-linear dynamic system with a linear non-stationary one using the first order Taylor expansion around the state estimates. This might imply that the extended Kalman smoother (EKS) is a better candidate --- in terms of implementation --- for the state inference in the proposed algorithm. However, the EKS requires re-linearising about the backward estimates which in turn necessitates the non-linear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation. In addition, the URTSS provides better state estimations compared to the EKS \cite{Haykin2001} and therefore the constructed lower bound will be more accurate, resulting in a better parameter estimation.

@@

Since the non-linearity is a sigmoid function first-order Taylor linearisation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated non-linear behaviour higher order terms might need to be included, this however introduces intractable computational complexity.

\section{Conclusion}
In this chapter, a method has been developed to estimate a spatially continuous neural field model from electrophysiological data, where the kernel spatial support and the field disturbance characteristics were unknown to the estimator.  It was shown that by choosing the synaptic parameter and the observation noise variance within a plausible range these unknown quantities can be identified. This improves the complexity and the accuracy of the EM algorithm by reducing the number of unknown variables in the estimation process. In addition to the connectivity kernel weights and the synaptic parameter, the observation noise and disturbance variances were also identified. The empirical distributions of the estimated non-kernel parameters showed small bias with very small standard deviations.
